---
title: "tidy data"
output: github_document
---

```{r message = FALSE}
library(tidyverse)
library(readxl)
library(dplyr)
library(ggplot2)
library(plotly)
```

## Tidy Data
```{r}
#load data set
smoke_df_2020 =read_csv("data/adult20.csv")
smoke_df_2019 =read_csv("data/adult19.csv")
smoke_df_2018 =read_csv("data/adult18.csv")
smoke_df_2017 =read_csv("data/adult17.csv")
smoke_df_2016 =read_csv("data/adult16.csv")
clean_df = smoke_df_2020 %>% 
  #clean variable names
  janitor::clean_names()
```

#The smoking trend in the US over 5 years 
let's clean the dataset for observing the smoking trends in these five years 
lets select the relavant questions first and the following explain the details of the variables
smoking_status: current smoking stastus.

```{r}
#cleaning data for 2019 and 2020 
cleaning_for_smoke_trend= function(dataset){
  dataset%>% 
  janitor::clean_names()%>% 
  select(cignow_a, smkcigst_a,sex_a,hispallp_a, agep_a)%>% 
  rename(smoke_freq = cignow_a ,  smoking_status = smkcigst_a, sex = sex_a,race =hispallp_a , age =  agep_a)}
#select(cignow_a, smkcigst_a,smkage_a)
#reminder for later: smoking_age =smkage_a 
smoke_trend_2020 = cleaning_for_smoke_trend(smoke_df_2020)
smoke_trend_2019  = cleaning_for_smoke_trend(smoke_df_2019)

#cleaning data for 2018, 2017,2016 
#rewrite the function since they have a different variable code
cleaning_for_smoke_trend_2 = function(dataset){
  dataset%>% 
  janitor::clean_names()%>% 
  select(smkstat2,sex, age_p, racerpi2)%>% 
  rename(smoking_status =smkstat2, age = age_p, race= racerpi2)}

smoke_trend_2018 = cleaning_for_smoke_trend_2(smoke_df_2018)
smoke_trend_2017 = cleaning_for_smoke_trend_2(smoke_df_2017)
```



the number of people smoking 
```{r}
 finding_smoking_trend = function(dataframe){
   dataframe %>% 
  select(smoking_status)%>%
  mutate(smoking_status = recode(smoking_status,"1" = "current every day smoker", "2" ="current some day smoker",  "3"= "former smoker","4" = "never smoker"))%>% 
  count(smoking_status)%>%
  drop_na()}

#apply the function to know the summary of smoking status 
#can i use a map function here?
#have to add year and undergo pivot_longer
#problem encountered: temporarily there is no data available in 2016
smoke_trend_20 = smoke_trend_2020 %>% 
  finding_smoking_trend()%>%
  mutate(year =  "2020")

smoke_trend_19 = smoke_trend_2019 %>% 
  finding_smoking_trend()%>%
  mutate(year =  "2019")

 smoke_trend_18= smoke_trend_2018 %>% 
  finding_smoking_trend()%>%
  mutate(year =  "2018")

smoke_trend_17 = smoke_trend_2017 %>% 
  finding_smoking_trend()%>%
  mutate(year =  "2017")

smoke_trend_overall = bind_rows(smoke_trend_17,smoke_trend_18,smoke_trend_19,smoke_trend_20)%>%
  rename(ppl_sum = n)%>%
  relocate(year)

#lets create a ggpplot
# ***task： try it to make it interactive soon....
#Problem: Should I use rate instead....???
smoke_trend_overall_plot = smoke_trend_overall %>% 
  ggplot(aes(x = year, y = ppl_sum, group = smoking_status)) + 
  geom_point(aes(color = smoking_status))+ geom_line(aes(color = smoking_status))+ 
 ggtitle("Overall smoking trend of the U.S in recent 4 years according to NHIS ")+
  labs(y = "sum of people")

#ggplotly(smoke_trend_overall_plot)
```
After knowing the overall trend... let's observe the distribution of smoking. 
In order to know their distribution, we have to obtain the information of their age, race and sex. 
We can group by these categories and making a distribution plot.
```{r}
#function creating the same table to find the distribution
#first do some data cleaning
 finding_smoking_dis =  function(dataframe){
   dataframe %>% 
  select(smoking_status, race, age, sex)%>%
  filter(sex== "1" | sex== "2")%>%
  subset(race != "97"|race != "98"|race != "99") %>% 
  mutate(sex = recode(sex,"1" = "male", "2" ="female"), 
         race =recode(race, "1" = "hispanic ", "2" ="white" , "3"="black/african american", "4"= "asian ", "5" = "aian", "6" = "aian and any other group", "7" ="other races"),
         #since 1,2 represents current daily/occasional smokers whereas 3,4 is former smoker and never smoke respectively; we define 1,2 as smoker whereas 3,4 as non-smoker
         smoking_status = 
           case_when(smoking_status == "1"|smoking_status == "2"~ "smoker",
                     smoking_status == "3"|smoking_status == "4"~ "non-smoker"))}

df1= finding_smoking_dis(smoke_trend_2020)%>%
  mutate(year =  "2020")
df2 = finding_smoking_dis(smoke_trend_2019)%>%
  mutate(year =  "2019")
df3= finding_smoking_dis(smoke_trend_2018)%>%
  mutate(year =  "2018")
df4 = finding_smoking_dis(smoke_trend_2017)%>%
  mutate(year =  "2017")



#***making interactive plot with option clicking on different years 
#year 2020 smoking distribution
plot_df1 = df1%>%
filter(smoking_status =="smoker")%>%
  count(race,age,sex)%>%
  ggplot(aes(x =race, y =age,)) + 
  geom_violin(aes(fill = race), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")+facet_grid(.~sex)+ 
  labs(title=" Distribution of smoking among agegroup and race in year 2020")+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 
plot_df2 = df2%>%
filter(smoking_status =="smoker")%>%
  count(race,age,sex)%>%
  ggplot(aes(x =race, y =age)) + 
  geom_violin(aes(fill = race), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")+ facet_grid(.~sex)+ 
  labs(title=" Distribution of smoking among agegroup and race in year 2019") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

plot_df3=df3%>%
filter(smoking_status =="smoker")%>%
  count(race,age,sex)%>%
  ggplot(aes(x =race, y =age)) + 
  geom_violin(aes(fill = race), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")+
   facet_grid(.~sex)+ 
  labs(title=" Distribution of smoking among agegroup and race in year 2018") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


plot_df4 = df4%>%
  filter(smoking_status =="smoker")%>%
  count(race,age,sex)%>%
  ggplot(aes(x =race, y =age)) + 
  geom_violin(aes(fill = race), alpha = .5) + 
  stat_summary(fun = "median", color = "blue")+
  facet_grid(.~sex)+ labs(title=" Distribution of smoking among agegroup and race in year 2017") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

#ggplotly(plot_df4)
#ggplotly(plot_df3)
#ggplotly(plot_df2)
#ggplotly(plot_df1)
```

